Computing in the Field of Rationals

### 2012-06-25 rational_ext.tex
_

  This paper connects the line of work starting with
  Filinski\cite{Filinski:1989:DCI:648332.755574} on the duality of
  computation \cite{Curien:2000,DBLP:conf/rta/Wadler05} with work on
  information preserving computation \cite{infeffects,rc2011}.
  Andrzej Filinski's masters thesis
  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
  symmetry underlying computation in the \lcal. Filinski showed that
  values are dual to continuations and that functions are dual to
  delimited continuations, thereby suggesting that
  values-and-functions and continuations-and-delimited-continuations
  are mirror images of the same phenomena.

@@

  We present a computational model whose types are \emph{the field of
    rational numbers}. This computational model is derived
  systemmatically from the {{Pi}} by considering a symmetric notion of
  duality. Unlike the \lcal which shows a classical De Morgan duality,
  here we have two axis of dualization and hence two dualities -- an
  additive duality, namely negative types and a multiplicative
  duality, namely fractional types.

@@

Andrzej Filinski's masters thesis
\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
underlying computation in the \lcal. Filinski showed that values are
dual to continuations, functions are dual to delimited
continuations. Filinksi also shows a De Morgan style duality, by
treating value pairs as the dual of continuation sums and value sums
as dual to continuation pairs.

@@

@@

To show the right-to-left direction we proceed by induction of the
sequence of {{|-->*}} derivations. Again the case for primtive
isomorphisms follows trivially. Taking the case of {{c1+c2}} we are
required to show that given a sequence 
{{<c1+c2,v,C> |-->* {[c', v', C]} }} there is a derviation tree for
{{c1+c2 v |--> v'}}. In the case that {{v:v1+b2}} has the form
{{left~v1}}, this follows by observing that the given sequence must have the form
{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. 
For the strictly smaller inner subsequence by induction we have
{{c1~v1 |--> v1'}} which lets us complete the derivation of
{{c1+c2~(left v1) |--> left v1'}}. The {{right v2}} follows similarly.

@@

It is important to note that in a compact closed category, the
dualizing action, usually denoted $(-)^\asterix$, is a contravariant
functor. However {{PiEE*}} is not compact closed and {{1/(-)}} is not
a functor specifically because divsion is not defined on 0 sized
types.

@@

Information Effects, Filinksi, De Morgan

>>>>

  This paper connects the line of work starting with
  Filinski\cite{Filinski:1989:DCI:648332.755574} on the duality of
  computation \cite{Curien:2000,DBLP:conf/rta/Wadler05} with work on
  information preserving computation \cite{infeffects,rc2011}.
  Andrzej Filinski's Masters thesis
  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
  symmetry underlying computation in the \lcal. Filinski showed that
  values are dual to continuations and that functions are dual to
  delimited continuations, thereby suggesting that
  values-and-functions and continuations-and-delimited-continuations
  are mirror images of the same phenomena.

@@

  We present a computational model whose types are \emph{the field of
    rational numbers}. This computational model is derived
  systematically from the {{Pi}} by considering a symmetric notion of
  duality. Unlike the \lcal which shows a classical De Morgan duality,
  here we have two axis of dualization and hence two dualities -- an
  additive duality, namely negative types and a multiplicative
  duality, namely fractional types.

@@

Andrzej Filinski's Masters thesis
\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
underlying computation in the \lcal. Filinski showed that values are
dual to continuations, functions are dual to delimited
continuations. Filinski also shows a De Morgan style duality by
treating pairs of values as dual to sums (choice) of continuation and
sums of values as dual to pairs of continuations. He, furthermore,
introduces a symmetric extension of the $\lambda$-calculus in which
call-by-value is dual to call-by-name.

This view of symmetry between the value fragment of the language and
the continuation fragment was refined by Curien and Herbelin
\cite{Curien:2000} who introduce the sequent-style
$\mu\tilde{\mu}$-calculus, that exhibits symmetries between values and
continuations and between call-by-value and call-by-name. The calculus
includes the type $A-B$ which is the dual of implication, i.e., a
value of type $A-B$ is a context expecting a function of type $A
\rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation
of type $B$. This is to be contrasted with our interpretation of a
value of that type as \emph{either} a value of $A$ or a demand for a
value of type $B$. The duality between call-by-name and call-by-value
was further investigated by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. The
$\mu\tilde{\mu}$-calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives
no interpretation to the subtraction connective and like the original
symmetric calculus of Filinski, introduces a duality that relates sums
to products and vice-versa.

Previous work on \emph{information effects} established the logically
reversible computational model {{Pi}} (and its extension {{Pi0}})
wherein computation is information preserving. The computational model
is derived from the isomorphisms of finite-types, every computation
admits an adjoint that expresses the logical ``inverse'' of a
computation and is complete for combinational circuits. The term model
of {{Pi}} is a so-called commutative bimonoidal category, a term used
for the categorification of a commutative semiring.

The language {{Pi}} however lacked a natural representation for
first-class functions due to first-order nature of all of its
values. In the light of the work of duality of computation it was felt
that ``right model'' for expressing functions in {{Pi}} would also
express symmetries similar to those in the calculi of Filinski, Curien
and Herbelin and Wadler. The technical goal of this paper is to
present the duality of computation in the setting of {{Pi}}.

The extended language, which we call {{PiEE}}, does not have one De
Morgan style dualizing action $(\_)^\bot$, but two distinct dualizing
actions -- an additive duality characterized by negative types $-(\_)$
and a multiplicative duality characterized by fractional types
$1/(\_)$. These negative and fractional dualities must not confused
with the duality in the additive and multiplicative fragment of linear
logic with the dualizing action $(\-)^\bot$. Each dualizing operation
preserves the underlying connective {{-(a+b)<->(-a)+(-b)}} and
{{1/(a*b)<->1/a*1/b}}.

The types of {{PiEE}} thus correspond to the field of rational numbers
(with the important restriction of fields that there is no division by
zero). A direct consequence of this is that algebraic manipulations of
rationals can be given an operational interpretation and \emph{have
  computational content}.

Models of linear logic rely on *-autonomous categories which change
the monoidal tensor when dualizing. Compact closed categories preserve
the monoidal tensor on dualizing.  The term model of {{PiEE}}
corresponds to the \emph{categorification of a field}. The category is
compact closed over the additive monoid. In the multiplicative monoid,
since there is no division by zero, respects compact closed structure
only when a dual object is defined.

Say more: fractionals, entaglements, backtracking, superposition,
constrants, negative information. Negative types, backward information
flows, sets with negative types, iteration.. etc.

We present {{Pi}} in the big-step style presented in the previous
work, develop the categorical semantics of its term model and present
a small-step semantics useful for modeling fractionals and
negatives. We proceed to present the fractional fragment of the
language {{PiEE*}}, work out its properties and categorical
semantics. We further use the multiplicative {{trace}} to express a
SAT solver. We then present the negative fragment of the language
{{PiEE+}} and finally present the unified language {{PiEE}}.

%% Traditionally, dealing with fields in a categorical setting has proven
%% hard and hence we hope that the term model of {{PiEE}} will prove
%% useful. Further, it is as yet unclear why there are two different
%% dualities in the setting of {{Pi}}.

%% \paragraph*{Declarative Continuations.} 
%% In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
%% proposes that continuations are a \emph{declarative} concept. He,
%% furthermore, introduces a symmetric extension of the $\lambda$-calculus in
%% which call-by-value is dual to call-by-name and values are dual to
%% continuations. In more detail, the symmetric calculus contains a ``value''
%% fragment and a ``continuation'' fragment which are mirror images. Pairs and
%% sums are treated as duals in the sense that the ``value'' fragment includes
%% pairs whose mirror image in the ``continuation'' fragment are sums. In
%% contrast, our language includes pairs and sums in the value fragment and two
%% symmetries: one that maps the pairs to fractions and another that maps the
%% sums to subtractions.

@@

  \begin{multicols}{2}
\dgm{c1c2_par_sum}
\dgm{c1c2_par_times}
  \end{multicols}
\dgm{c1c2_seq}

@@

To show the right-to-left direction we proceed by induction of the
sequence of {{|-->*}} derivations. Again the case for primtive
isomorphisms follows trivially. Taking the case of {{c1+c2}} we are
required to show that given a sequence 
{{<c1+c2,v,C> |-->* {[c', v', C]} }} there is a derivation tree for
{{c1+c2 v |--> v'}}. In the case that {{v:v1+b2}} has the form
{{left~v1}}, this follows by observing that the given sequence must have the form
{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. 
For the strictly smaller inner subsequence by induction we have
{{c1~v1 |--> v1'}} which lets us complete the derivation of
{{c1+c2~(left v1) |--> left v1'}}. The {{right v2}} follows similarly.

@@

It is important to note that in a compact closed category, the
dualizing action, usually denoted $(-)^\asterix$, is a contravariant
functor. However {{PiEE*}} is not compact closed and {{1/(-)}} is not
a functor specifically because division is not defined on 0 sized
types.

@@

Information Effects, Filinski, De Morgan
